DBES Grand Challenge : High throughput, scalable solution for processing taxi trip data

### 2015-03-28 paper/introduction.tex
updating the introduction

\section{Introduction}
Stream processing systems are particularly suitable for processing tuples that may be reported by devices, computing processes, or humans. These streams are processed to perform knowledge extraction based on the information encoded within individual tuples. Given the rate at which these tuples are generated, inability of the processing to keep pace with the data generation rates will lead to buffer overflows, dropped packets, and incorrect results. Often the data from these streams must be processed in real time and the knowledge based updated as well.

In this paper, we provide our solution to the DEBS Grand Challenge problem of processing New York taxi trip data and to perform knowledge extraction. We consider this problem in the context of stream processing where the tuples are passed through stream processing stages and the knowledge base updated in real time.
\subsection{Research Challenges}
Processing trip data streams and continually updating the knowledge base includes the following challenges:
\begin{enumerate}
	\item \textit{Data is continually arriving.} The processing of the data streams must keep pace with the data generation rates.
	\item \textit{Regardless of the data arrival rates, at any given instant the knowledge base must be accurate and current:} Even as tuples continually arrive, the information desired by the system (the top 10 values) must reflect the current operating conditions. It is possible that as a new trip information arrives, the list of top 10 routes may change. This update must occur in real time as soon as the tuple that triggered this change is processed.
	\item \textit{Updates to data structures must be accurate and keep pace with data arrival rates.} The time and space complexity of the data structures used to manage the knowledge base must be taken into account. Specifically, the data structure that we use must not preclude real time processing.
\end{enumerate}

\subsection{Research Questions}
We consider several research questions in our solution to the DEBS Grand Challenge problem. These include:
\begin{enumerate}
	\item \textit{How can we support real-time analytics and processing over these streams?}  Processing at this scale and in real-time crosscuts algorithmic and system design. The framework must be able to scale with increase in data volumes and arrival rates.
	\item \textit{How can we extract and update knowledge base in real time?}  Since the data streams are continually arriving the space and time complexity of the data structures become important. The data structures must be compact to conserve memory and the time-complexity of the operations to update the structures must be low.
	\item \textit{How can we ensure high-throughput processing to support processing a large number of packets per second?} This could involve leveraging multiple instances of various processing stages. 
\end{enumerate}

>>>>

\section{Introduction}

The Grand Challenge 2015 problem focuses on processing New York city taxi trip data for year 2013. The area considered is a square of 150m x  150m. The  square is divided into smaller units called cells and a taxi trip is considered as moving from one cell (pickup cell) to another cell (dropoff cell). Therefore a route is uniquely identified by the pickup cell and the dropoff cell. The data set consists of 173 million events and solutions are expected to replay these data and generate events that meet criteria of two queries defined.

The first query is to find the events which change the top ten routes within last 30 minutes and emit the new route set details along the original event. The given data set contains 899010  routes and the 30 minute time window for this query accounts around 10000 events assuming an equal event distribution. This creates the possibility of processing several thousand route counts specially within peak ours. In other words we need to evaluate several thousand route counts with each event and find 10 top most routes. Similar analysis can be made with profitable cells query as well. In addition to that it requires to  calculate the median fare for each cell.

\subsection{Research Challengers}

If we consider an application to process this 173 million data set within an year, it would be a trivial task and any naive implementation satisfy that requirement. However if we consider an application which process these data within minutes, we observe following research challengers.
\begin{enumerate}
	\item \textit{Handling thousands of values in real time :}  An algorithm with O(n) time complexity will not scale for thousands. Therefore is it possible to come up with an algorithm of O(log(n)) time complexity?
	\item \textit{Parallel evaluation :}  This application is not inherently support parallel evaluation. Partitioning route sets and cell sets triggers events for sub sets, but application requires the events with whole sets. Therefore is it possible to partition data and still generate required results to support parallelism?  
\end{enumerate}

\subsection{Methodology}

We started our work seeking answers for above challengers. First we evaluated doubly linked list and heap based approaches to build a data structure with O(log(n)) time complexity. After several evaluations, we developed a new data structure called \textit{NodeList} which utilizes both concepts. In order to find the running median with O(log(n)) time complexity, we experimented with an algorithm which uses priority queue and the heap data structure we developed. We decided to use former since it gave better performance with smaller window sizes. Then we focused on parallelizing the application and came up with an strategy to aggregate the sub set results to find the final results. As the last step, we conducted several experiments to evaluate our solutions. We repeated this process identifying possible issues and fixing them.
